Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2022-01-31 document/source/body.tex
GC KE meeting

Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure, failing to detect bugs and potentially leading to false confidence in the system under test. When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions are:
{\em How can one assess whether a simulation environment is deterministic?} and 
{\em How can one determine and control the simulation variance?}

@@

The game loop cycle starts with initialising the scene and actors. Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate. This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}. The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}. When the frame is rendered the game loop cycle returns to processing inputs. An intuitive and  more detailed description of the interplay between the physics and render cycles, see\footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given in~\cite{JohnAustinUnity}.

@@

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which is limited by the fixed bit width available and can potentially result rounding errors~\cite{FloatingPointsBook,goldberg1991every}. As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

@@

It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output. However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
% ~\cite{unity_multithreading} 
or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.

>>>>

Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may render the outcome unstable producing incorrect results potentially leading false confidence in the system under test. 
%in intermittent assertion failure, failing to detect bugs and potentially leading to false confidence in the system under test. 
When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions are:
{\em How can one assess whether a simulation environment is deterministic?} and 
{\em How can one determine and control the simulation variance?}

@@

The game loop cycle starts with initialising the scene and actors. Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate. This is illustrated by the loop in the physics update in Fig.~\ref{GameEngineLoopDiagram}. The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate~\cite{unity_framerates}. When the frame is rendered the game loop cycle returns to processing inputs. For an intuitive and  more detailed description of the interplay between the physics and render cycles, see~\cite{JohnAustinUnity}. 
% \footnote{\url{https://johnaustin.io/articles/2019/fix-your-unity-timestep}}.% is given

@@

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which necessitates rounding due to a fixed bit width~\cite{FloatingPointsBook,goldberg1991every}. 
% available and can potentially result in rounding errors~\cite{FloatingPointsBook,goldberg1991every}. 
As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and results may differ depending on execution order.
% Some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, this
% finite-precision of floating point number representation 
% such rounding errors 
could result in accuracy issues of, for example, actor positions. While some authors suggest
avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and performing the execution on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

@@

It is responsibility of the tester
% It is unlikely 
It's unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output. However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
% ~\cite{unity_multithreading} 
or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.
